Predicting complex diseases: performance and robustness [OF NEW METHOD?]

### 2018-03-01 5-results-celiac.Rmd
add celiac results

---
title: "Analysis of celiac disease"
author: "Florian Privé"
date: "November 3, 2017"
output: html_document
---

@@

```{r}
# Read file, see "Preprocessing" notebook
library(bigsnpr)
celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- nb_cores()
y <- celiac$fam$affection - 1
```

@@

```{r}
# Divide in training/test sets
dim(G)
ind.train <- sort(sample(nrow(G), size = 12e3))
ind.test <- setdiff(rows_along(G), ind.train)
```

@@

```{r}
# Q-Q plot
snp_qq(gwas.train) +
  coord_cartesian(ylim = c(0, 20))
gwas.train.gc <- snp_gc(gwas.train)
```

```{r, dev="png"}
# Manhattan plot
labels <- 1:22; labels[c(11, 13, 15, 17, 18, 20, 21)] <- ""
cowplot::plot_grid(
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels),
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels) +
    coord_cartesian(ylim = c(0, 22)) + 
    geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
  align = "hv", ncol = 1, labels = LETTERS[1:2], label_size = 25, scale = 0.95
)
```

@@

```{r}
# PRS
thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
lpS <- -predict(gwas.train.gc)
prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
               ind.test = ind.test,
               ind.keep = ind.keep,
               lpS.keep = lpS[ind.keep], 
               thr.list = thrs)
nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
aucs <- apply(prs, 2, AUC, target = y[ind.test])
summary(aucs)
```

```{r}
pAUC <- function(pred, target, p = 0.1) {
  val.min <- min(target)
  q <- quantile(pred[target == val.min], probs = 1 - p)
  ind <- (target != val.min) | (pred > q)
  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
           error = function(e) p^2 / 2)  # all same prediction
}
plot_density_AUC <- function(pred, true = y[ind.test]) {
  data.frame(pred = pred, pheno = as.factor(true)) %>%
  myggplot() +
  geom_density(aes(pred, fill = pheno), alpha = 0.4) +
  ggtitle(sprintf("AUC = %s%% | pAUC = %s", 
                  round(100 * AUC(pred, true), 2),
                  round(pAUC(pred, true), 4)))
}
plot_density_AUC(prs[, which.max(aucs)])
```

@@

```{r}
plot_density_AUC(preds1.comb)
```

@@

```{r}
plot_density_AUC(preds2.comb)
```

### Regularization paths

```{r}
h <- round(c(max(aucs), auc.cmsa, auc.cmsa2), 4)
p <- myggplot(rbind(
  cbind.data.frame(nb = nb.pred,      AUC = aucs,      Method = "PRS"),
  cbind.data.frame(nb = nb.pred1[-1], AUC = aucs1[-1], Method = "logit-simple"),
  cbind.data.frame(nb = nb.pred2[-1], AUC = aucs2[-1], Method = "logit-triple")
)) +
  geom_point(aes(nb, AUC, color = Method)) +
  scale_x_log10(breaks = 10^(0:6)) +
  labs(x = "Number of predictors") +
  geom_hline(yintercept = h, color = scales::hue_pal()(3), linetype = 2) +
  theme(legend.position = c(0.85, 0.5))
p + 
  scale_y_continuous(breaks = sort(c(ggplot_build(p)$layout$panel_ranges[[1]]$y.major_source, h)))
```

```{r}
ggsave("figures/main-celiac-regpath.pdf", scale = 1/90, width = 1000, height = 700)
```

@@

```{r, fig.asp=1, fig.width=8}
library(plotROC)
scores_tidy <- tibble(
  d = y[ind.test], 
  "PRS-max" =  prs[, which.max(aucs)],
  "logit-simple" = preds2,
  "logit-triple" = preds4
) %>%
  gather(key = "Method", value = "Score", -d)

@@

```{r}
ggsave("figures/supp-celiac-roc.pdf", scale = 1/90, width = 700, height = 700)
```

@@

```{r}
theone <- which.max(abs(gwas.train.gc$score))
ind0 <- which(G[ind.test, theone] == 0)
(M <- length(ind0))
table(Geno = G[, theone], forcats::fct_recode(as.factor(y), Control = "0", Case = "1"))
```

```{r}
# one geno
mean(y[ind.test[ind0]]) 
# prs max
indPRS <- order(prs[, which.max(aucs)], decreasing = TRUE)[1:M]
mean(y[ind.test[indPRS]])
# logit-simple (CMSA)
indLS <- order(preds2, decreasing = TRUE)[1:M]
mean(y[ind.test[indLS]])
# logit-triple (CMSA)
indLT <- order(preds4, decreasing = TRUE)[1:M]
mean(y[ind.test[indLT]])
```

## Stability of selection of the "logit-simple"

Train on 100 bootstrap samples of the Celiac dataset.

```{r, eval=FALSE}
if (!dir.exists("results6")) dir.create("results6")

for (i in seq_len(100)) {

  res.file <- paste0("results6/simu_", i, ".rds")
  if (file.exists(res.file)) next

  ind.train <- sort(sample(nrow(G), replace = TRUE))

  cmsa.logit <- big_CMSA(FUN = big_spLogReg, feval = AUC,
                         X = G, y.train = y[ind.train], 
                         ind.train = ind.train, 
                         covar.train = obj.svd$u[ind.train, ],
                         alpha = 0.5, dfmax = 20e3, 
                         ncores = NCORES)

  saveRDS(list(betas = cmsa.logit[cols_along(G)],
               ind.test = setdiff(rows_along(G), ind.train)),
          file = res.file)
}
```

```{r}
results6 <- list.files("results6", full.names = TRUE) %>%
  map(~readRDS(.x)) %>%
  do.call("cbind", .)
```

```{r}
n_enter <- rowSums(results6 != 0)
mean(n_enter == 0)                ## Proportion of SNPs never entering
n_enter_once <- n_enter[n_enter != 0]
mean(n_enter_once)                ## Mean number of times a SNPs enter 
                                  ## if entering at least once
bigstatsr:::MY_THEME(
  qplot(n_enter_once, geom = "bar") +
    labs(x = "Number of times")
)
```

```{r}
ggsave("figures/supp-celiac-stabilitybar.pdf", scale = 1/90, width = 1000, height = 700)
```

```{r}
length(always_enter <- which(n_enter == ncol(results6)))
cowplot::plot_grid(
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
                ind.highlight = always_enter),
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels, 
                ind.highlight = always_enter) +
    coord_cartesian(ylim = c(0, 25)) + 
    geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
  align = "hv", ncol = 1, labels = LETTERS[1:2], label_size = 25, scale = 0.95
)
```

```{r}
ggsave("figures/supp-celiac-man-col.png", scale = 1/90, width = 1300, height = 1000)
```

```{r}
sq_effects_always <- colSums((results6[always_enter, ])^2)
sq_effects_others <- colSums((results6[-always_enter, ])^2)
summary(100 * sq_effects_always / (sq_effects_always + sq_effects_others))
```

>>>>

---
title: "Analysis of celiac disease"
author: "Florian Privé"
date: "March 1, 2018"
output: html_document
---

@@

```{r, cache=FALSE}
# Read file, see "Preprocessing" notebook
library(bigsnpr)
celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- nb_cores()
y <- celiac$fam$affection - 1
```

@@

```{r}
# Divide in training/test sets
dim(G)
set.seed(1)
ind.train <- sort(sample(nrow(G), size = 12e3))
ind.test <- setdiff(rows_along(G), ind.train)
```

@@

```{r, dev='png', fig.asp=0.8}
# Q-Q plot
snp_qq(gwas.train) +
  coord_cartesian(ylim = c(0, 15))
gwas.train.gc <- snp_gc(gwas.train)
```

```{r, dev="png", fig.asp=0.9, out.width="90%"}
# Manhattan plot
labels <- 1:22; labels[c(11, 13, 15, 17, 18, 20, 21)] <- ""
cowplot::plot_grid(
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels),
  snp_manhattan(gwas.train.gc, CHR, POS, labels = labels) +
    coord_cartesian(ylim = c(0, 22)) + 
    geom_hline(yintercept = -log10(5e-8), color = "red", linetype = 3),
  align = "hv", ncol = 1, labels = LETTERS[1:2], label_size = 25, scale = 0.95
)
```

@@

```{r}
# PRS
thrs <- c(0, -log10(5e-8), exp(seq(log(0.1), log(100), length.out = 100)))
lpS <- -predict(gwas.train.gc)
prs <- snp_PRS(G, betas.keep = gwas.train.gc$estim[ind.keep],
               ind.test = ind.test,
               ind.keep = ind.keep,
               lpS.keep = lpS[ind.keep], 
               thr.list = thrs)
nb.pred <- sapply(thrs, function(thr) sum(lpS[ind.keep] > thr))
aucs <- apply(prs, 2, AUC, target = y[ind.test])
summary(aucs)
nb.pred[which.max(aucs)]
```

```{r}
pAUC <- function(pred, target, p = 0.1) {
  val.min <- min(target)
  q <- quantile(pred[target == val.min], probs = 1 - p)
  ind <- (target != val.min) | (pred > q)
  tryCatch(bigstatsr::AUC(pred[ind], target[ind]) * p, 
           error = function(e) p^2 / 2)  # all same prediction
}
plot_density_AUC <- function(pred, true = y[ind.test]) {
  data.frame(pred = pred, pheno = as.factor(true)) %>%
  myggplot() +
  geom_density(aes(pred, fill = pheno), alpha = 0.4) +
  ggtitle(sprintf("AUC = %s%% | pAUC = %s", 
                  round(100 * AUC(pred, true), 2),
                  round(pAUC(pred, true), 4))) + 
  theme(plot.title = element_text(size = rel(1.5), hjust = 0.5))
}
(p1 <- plot_density_AUC(prs[, which.max(aucs)]))
```

@@

```{r}
(p2 <- plot_density_AUC(preds1.comb))
```

@@

```{r}
(p3 <- plot_density_AUC(preds2.comb))
```

```{r, out.width="70%", fig.asp=900/780}
list(p1, p2, p3) %>%
  lapply(function(p) p + theme(legend.position = "none")) %>%
  cowplot::plot_grid(plotlist = ., ncol = 1, labels = LETTERS[1:3], label_size = 20, scale = 0.95) %>%
  cowplot::plot_grid(., cowplot::get_legend(p1), rel_widths = c(1, 0.15), scale = 0.95)
ggsave("figures/supp-score-densities.pdf", scale = 1/100, width = 780, height = 900)
```

@@

```{r, fig.asp=1, fig.width=8}
library(plotROC)
scores_tidy <- tibble(
  d = y[ind.test], 
  "PRS-max"      = prs[, which.max(aucs)],
  "logit-simple" = preds1.comb,
  "logit-triple" = preds2.comb
) %>%
  gather(key = "Method", value = "Score", -d)

@@

```{r}
ggsave("figures/celiac-roc.pdf", scale = 1/90, width = 700, height = 700)
```

@@

```{r}
# Most significant SNP
theone <- which.max(-predict(gwas.train.gc))
ind0 <- which(G[ind.test, theone] == 0)
(M <- length(ind0))
table(Geno = G[, theone], forcats::fct_recode(as.factor(y), Control = "0", Case = "1"))
```

```{r}
# one geno
mean(y[ind.test[ind0]]) 
# prs max
indPRS <- order(prs[, which.max(aucs)], decreasing = TRUE)[1:M]
mean(y[ind.test[indPRS]])
# logit-simple (CMSA)
indLS <- order(preds1.comb, decreasing = TRUE)[1:M]
mean(y[ind.test[indLS]])
# logit-triple (CMSA)
indLT <- order(preds2.comb, decreasing = TRUE)[1:M]
mean(y[ind.test[indLT]])
```

### 2018-03-01 paper/draft.tex
add celiac results

\title{Predicting complex diseases: performance, robustness and ..utility..(impact?)}
\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}

@@

\noindent$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France \\
\noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.

@@

We recently developed two R packages, bigstatsr and bigsnpr, for efficient analysis of large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr includes efficient functions for computing sparse linear and logistic regressions on huge datasets. In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}. 
Note that in this comparison we don't include any LMM method for the reasons mentioned before and don't include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
For the C+T model, we compare different thresholds of inclusion of SNPs.
For the logistic regression, we include two novel approaches. 
First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the amount of regularization used (one hyper-parameter of the penalized logistic regression), which directly affects the number of SNPs included in the model. 
We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.

@@

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in our package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values). To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is chosen. Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ using the geometric median) or resulting scores can be combined (e.g.\ averaging the $K$ probabilities for each individual). We also use this procedure to derive an early stopping criterion so that the algorithm doesn't need to evaluate the whole regularization path, making the whole procedure much faster.
Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We call this method ``logit-simple'' in the results.

@@

\subsection{Overview and naming}

In results, we use estimates of maximum achievable AUC values for the simulations: 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%) based on three estimations (see Supplementary Materials).

@@

\subsection{[Other measures of performance]}

Our method is parallelized and requires only approximately 500 seconds to run (with 6 physical cores) for a training size of 6000 individuals. It takes 1000 seconds to train on 12,000 individuals. Moreover, when there are 3 times more variables, it takes approximately 3 more times to run. We can confidently say that the ``logit-simple'' is of linear complexity in both dimensions.
Moreover, the ``logit-simple'' results in relatively sparse models with no more than 14,000 predictors entering the models even for the linear model of 3000 causal SNPs (Figure \ref{fig:supp-triple}). The ``logit-triple'' is comparable in terms of number of predictors.
Finally, for other predictive performance measures, the percentage of cases in either the 10\% or 20\% largest PRSs, we report that their values are correlated to the values of AUCs at more than 98\%, which is why we don't speak about them more in the results (Figure \ref{fig:supp-AUC-corr}).

@@

[MONTRER UN MANHATTAN PLOT POUR JUSTIFIER 30 IN HLA]

We also analyze the whole Celiac dataset with the original real phenotypes. For one run, splitting the dataset in 12,000 indiduals in the training set and the remaining 3155 individuals in the test set, we compare the ``regularization paths'' of all methods (T-Trees excepted), that is to say the predictive performance as a function of the number of predictors corresponding to different inclusion thresholds of the methods. 
We show that the regularization paths of both logistic regressions are both very smooth, reaching a maximum AUC or more than 89\% at approximately 1000 predictors (Figure \ref{fig:main-celiac-regpath}). Moreover, the CMSA procedure give AUC very close to the maximum AUC of the regularization path. For the C+T method, on the contrary, the regularization path is not smooth at all and the AUC never exceed the AUC given by the most stringent threshold [ADD VALUE OF AUC] ($p < 10^{-100}$). We also show some associated ROC curves [QUOI DIRE DE CES ROC?] (Figure \ref{fig:supp-celiac-roc}).

@@

In this comparative study, we propose other methods as a replacement to the widely used C+T method. 
We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method.
We are well aware that the C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet. 
However, in disease prediction for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only [EVEN LARGE CONSORTIA DON'T HAVE ALL DATA?].

@@

@@

\newpage
\begin{table}[htbp]
\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
\vspace*{0.5em}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{simple} & PRS \\
& & & 30 in all & & & & logit-simple \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{fancy} & logit-triple \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & PRS \\ 
& & & & & & & logit-simple\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-celiac-regpath}}
\caption{``Regularization paths'' for the C+T, ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals. For the C+T method, the dotted line represents the maximum AUC while for the logistic regressions, the dotted lines represent the value using the CMSA procedure to be free of the choice of the threshold.}
\label{fig:main-celiac-regpath}
\end{figure}

@@

Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01). We are also grateful to F\'elix Balazard for useful discussions about T-Trees.

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
\caption{Scenario \textnumero1. Comparison of different predictive measures. Correlation is greater than 98\%.[ADD]}
\label{fig:supp-AUC-corr}
\end{figure}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-celiac-roc}}
\caption{ROC Curves for the ``C+T'', ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals.}
\label{fig:supp-celiac-roc}
\end{figure}

>>>>

\title{Predicting complex diseases: performance and robustness [OF NEW METHOD?]}
\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}

@@

\noindent$^{\text{\sf 1}}$Universit\'e Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Int\'egrative (C3BI), Institut Pasteur, Paris, France.

@@

We recently developed two R packages, bigstatsr and bigsnpr, for efficient analysis of large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr includes efficient functions for computing sparse linear and logistic regressions on huge datasets. In this paper, we present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in \cite{botta2014exploiting}. 
Note that in this comparison we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}.
For the C+T model, we compare different thresholds of inclusion of SNPs.
For the logistic regression, we include two novel approaches. 
First, we introduce a procedure that we call Cross-Model Selection and Averaging (CMSA) for automatically choosing the amount of regularization used (one hyper-parameter of the penalized logistic regression), which directly affects the number of SNPs included in the model. 
We also show how to use feature engineering in order to capture not only linear effects, but also recessive and dominant effects.

@@

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in our package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values). To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is chosen. Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ using the geometric median) or resulting scores can be combined (e.g.\ averaging the $K$ probabilities for each individual). We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization path, making the whole procedure much faster.
Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We call this method ``logit-simple'' in the results.

@@

\subsection{Maximum AUCs}

In results, we use estimates of maximum achievable AUC values for the simulations: 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%) based on three estimations (see Supplementary Materials). Note that we do not report the simulation results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).

@@

@@

We also analyze the whole Celiac dataset with the ..original/real.. phenotypes. For one run [ENOUGH FOR REVIEWERS?], splitting the dataset in 12,000 indiduals in the training set and the remaining 3155 individuals in the test set, we compare the predictive performance of the ``C+T'', ``logit-simple'' and ``logit-triple'' methods, but also computation times and number of predictors used in the models (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
We ..see.. that the joint logistic regressions provide higher AUC with 88.71\% with the ``logit-simple'' and 89.26 with the ``logit-triple as compared to 82.46 with the ``C+T'' method, and much higher partial AUC (specificity 0.9-1), respectively 0.0411, 0.0440 and 0.0283, while using less predictors, respectively 2198, 4617 and 13277. Note that for the ``C+T'' method, we still report the best result among 102 p-values thresholds. In terms of computation time, we report only the GWAS computation for the ``C+T'' method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the fast ``C+T'' method, and the ``logit-triple'' takes twice as long as the ``logit-simple''. 
Finally, note that the scores you get with the logistic regressions are probabilities, which is not the case with the ``C+T'' method (Figure \ref{fig:supp-score-densities}).[QUE RAPPORTER SUR LES DENSITY PLOTS DU SUPMAT?]

\begin{table}[h]
\caption{Results for the real Celiac dataset.\label{tab:results-celiac}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
Method & AUC (\%) & pAUC & \# predictors & Execution time (s) \\ \hline
PRS-max & 82.46 & 0.0283 & 13277 & 149 \\ 
logit-simple & 88.71 & 0.0411 & 2198 & 78 \\
logit-triple & 89.26 & 0.0440 & 4617 & 161 \\ \hline
\end{tabular}
\end{table}

%We show that the regularization paths of both logistic regressions are both very smooth, reaching a maximum AUC or more than 89\% at approximately 1000 predictors (Figure \ref{fig:main-celiac-regpath}). Moreover, the CMSA procedure give AUC very close to the maximum AUC of the regularization path. For the C+T method, on the contrary, the regularization path is not smooth at all and the AUC never exceed the AUC given by the most stringent threshold [ADD VALUE OF AUC] ($p < 10^{-100}$). We also show some associated ROC curves [QUOI DIRE DE CES ROC?] (Figure \ref{fig:celiac-roc}).

@@

In this comparative study, we propose other methods as a replacement to the widely used C+T method. 
We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method that uses SNP effects learned independently.
We are well aware that the C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet. 
However, in disease prediction for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only [EVEN LARGE CONSORTIA DON'T HAVE ALL DATA?]. The logistic regression we present here could be used on data as large as biobank datasets because it is very efficient in both memory and computation time (..even faster than doing a GWAS!..).

@@

We also introduce a simple feature engineering technique to detect and learn dominant and recessive effects..

@@

\newpage
\begin{table}[htbp]
\caption{Summary of all simulations. Where there is symbol `-' in a box, it means that the parameters are the same as the ones in the upper box.\label{tab:simus}}
\vspace*{0.5em}
\centering
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
Numero of & \multirow{2}{*}{Dataset} & Size of & Causal SNPs & Distribution & \multirow{2}{*}{Heritability} & Simulation & \multirow{2}{*}{Methods} \\ 
scenario & & training set & (number and location) & of effects & & model & \\
\hline
\hline
\multirow{4}{*}{1} & \multirow{4}{*}{All 22 chromosomes} & \multirow{4}{*}{6000} & 30 in HLA & \multirow{2.5}{*}{Gaussian} & \multirow{2.5}{*}{0.5} & \multirow{2.5}{*}{simple} & PRS \\
& & & 30 in all & & & & logit-simple \\
& & & 300 in all & \multirow{1.5}{*}{Laplace} & \multirow{1.5}{*}{0.8} & \multirow{1.5}{*}{fancy} & logit-triple \\
& & & 3000 in all & & & & (T-Trees) \\
\hline
\multirow{2}{*}{2} & \multirow{2}{*}{Chromosome 6 only} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{-} & \multirow{2}{*}{simple} & PRS \\ 
& & & & & & & logit-simple\\
\hline
\multirow{5}{*}{3} & \multirow{5}{*}{All 22 chromosomes} & 1000 & \multirow{5}{*}{300 in all} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} & \multirow{5}{*}{-} \\ 
& & 2000 & & & & & \\
& & 3000 & & & & & \\
& & 4000 & & & & & \\
& & 5000 & & & & & \\
\hline
\end{tabular}
\end{table}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{celiac-roc}}
\caption{ROC Curves for the ``C+T'', ``logit-simple'' and ``logit-triple'' methods for one run on the Celiac disease, training on 12,000 individuals and projecting on the remaining 3155 individuals. Plotted using R package plotROC \cite[]{sachs2017plotroc}.\label{fig:celiac-roc}}
\end{figure}

@@

Authors acknowledge Grenoble Alpes Data Institute, supported by the French National Research Agency under the ``Investissements d'avenir'' program (ANR-15-IDEX-02) and the LabEx PERSYVAL-Lab (ANR-11-LABX-0025-01). We are also grateful to F\'elix Balazard for useful discussions about T-Trees, and to Yaohui Zeng for useful discussions about R package biglasso.

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-corr}}
\caption{Scenario \textnumero1. Comparison of different predictive measures. ..There is a Spearman correlation.. of 98\% between values of AUC and partial AUC.[ADD]\label{fig:supp-AUC-corr}}
\end{figure}

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-score-densities}}
\caption{[ADD]\label{fig:supp-score-densities}}
\end{figure}
